Containers and R: the Rockerverse and beyond

### 2019-08-19 manuscript.Rmd
Adding sections about
- Future

- bioconductor_full

- Cloud activities

- recent improvements to bioc_docker (audit and optimization)

@@

_Bioconductor_ is a free, open source and open development project for the analysis and comprehension of genomic data [1].
The project consists of 1741 R software packages, as well as packages containing annotation or experiment data.
_Bioconductor_ has a semiannual release cycle, each release is associated with a particular version of R.
Docker images allow availability of current and past versions of _Bioconductor_ for convenience and reproducibility.

_Bioconductor_ docker images are built on top of rocker/r-ver and rocker/rstudio.
By using rocker as a base image, the R version and most package and system dependencies are included in a _Bioconductor_ base image.
_Bioconductor_ installs packages based on the R version, and therefore leverage the rocker/rstudio and rocker/r-ver use of R version tagging.
_Bioconductor_ selects the desired version of R from rocker, adds the BiocManager CRAN package for installing appropriate versions of _Bioconductor_ packages, and creates a _Bioconductor_ docker image with a informative tag (R\_version\_Bioc\_version).
The images are summarized on the _Bioconductor_ web site [5], maintained in github [6], and available to the community through DockerHub [3].
All past and current combinations of R and _Bioconductor_ are therefore accessible via a specific docker tag.

@@

@@

>>>>

	- kubernetes orchestrated

@@

_Bioconductor_ is a free, open source and open development project for
the analysis and comprehension of genomic data [1]. The project
consists of 1741 R software packages as of August 15th 2019, as well
as packages containing annotation or experiment data. _Bioconductor_
has a semiannual release cycle, each release is associated with a
particular version of R. Docker images allow availability of current
and past versions of _Bioconductor_ for convenience and
reproducibility.

_Bioconductor_ docker images are built on top of rocker/r-ver and
rocker/rstudio. By using rocker as a base image, the R version and
most package and system dependencies are included in a _Bioconductor_
base image.  _Bioconductor_ installs packages based on the R version,
and therefore leverage the rocker/rstudio and rocker/r-ver use of R
version tagging. _Bioconductor_ selects the desired version of R from
rocker, adds the BiocManager CRAN package for installing appropriate
versions of _Bioconductor_ packages, and creates a _Bioconductor_
docker image with an informative tag (R\_version\_Bioc\_version). The
images are summarized on the _Bioconductor_ web site [5], maintained
in github [6], and available to the community through DockerHub [3].
All past and current combinations of R and _Bioconductor_ are
therefore accessible via a specific docker tag.

@@

Bioconductor's _bioc_docker_ software works using a templating engine
based in ruby, has a source directory `src` and an output directory
`out`.[*] This takes away in large part the repeated handcrafting of
each Dockerfile produced by Bioconductor at each release cycle. Edits
to the Dockerfile need to be made in the source directory, based on a
config file, the output directory is populated with the appropriate
Dockerfile for each specific to the various areas of research. A
recent systematic "audit" of our docker images following best
practices (as issued by the Docker website) led to a reduction in the
number of layers and also size of the overall images being produced by
Bioconductor. Some of the optimizations include removing cache,
cleaning up after each layer produced by a `RUN` statement, and most
importantly, not re-installing any dependencies already provided by
rocker.

Bioconductor now produces a Docker image to mimic the _Bioconductor_
nightly build system on Linux. This image is called the
`bioconductor_full` (based on rocker/rstudio). The goal of the Docker
image is to install _all the system dependencies_ needed to install
all of the Bioconductor packages. We achieve this goal to a large
extent, and are able to greater than _99%_ of packages (1730 out of
1741). This image allows developers to test their packages extensively
by replicating errors on the nightly build system, and users to
install most of the packages they need without dealing with system
dependencies. There is the added bonus that it comes with an RStudio
interface to ease users into using R with Docker, thanks to rocker.

We found this to be a easier alternative to distributing installed
_Bioconductor_ packages within Docker images as this is not entirely
consistent with the way R/Bioconductor is used. Users usually install
and add packages to their scripts in real time. It also keeps the
Docker image smaller in size. Using R is almost never like a
'workflow', which is defined by the pre-installed packages. We
encourage users to mount a volume to their Bioconductor docker image,
where all the installed (compiled) packages are stored in the external
mountable volume. The library tree `.libPaths()` has
`/usr/local/lib/R/host-site-library` as the first location searched
for installed packages, which allows this functionality to work. This
enables the docker image to be updated by using a `docker pull`
command without the user having to worry about any packages saved on
the image.

The `bioconductor_full` image is forming the base for our cloud based
solutions in upcoming future projects. We are currently able to host a
Kubernetes orchestrated cluster on the Google Cloud Platform using the
bioconductor_full image and the rocker/r-ver images. The distribution
of computation is done through an R package called `RedisParam`
(github: mtmorgan/RedisParam), where the `bioconductor_full` image is
the master node and the `rocker/r-ver` images are the workers.

We envision these Docker images being the primary way of teaching in
the future. The goal is to let workshop participants pull a pre-built
docker image for the specific worksop and run them on their local
machines. As of right now, the same task is accomplished by creating
and spinning up multiple AWS images (one per participant). The docker
images can work in more fault tolerant, lasting, cost-effective
manner, and be used by participants throughout their learning
journey. The users can choose to scale their workshop skills using
their cloud accounts to scale up with this provided docker image
(using either the AWS Elastic Container service or Google Kubernetes
Engine etc.)

@@

[*] The original work on `bioc_docker` was done by Dan Tanenbaum
